---
title: 'Assignment 10: Data Scraping'
author: "Wynona Curaming"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse);library(lubridate);library(viridis);library(here);library(rvest)
library(dataRetrieval);library(tidycensus);library(ggplot2);library(dplyr)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
Water_URL<- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
Water_URL
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system_name<-Water_URL%>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)")%>%
  html_text()
water_system_name

PWSID<-Water_URL%>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)")%>%
  html_text()
PWSID

ownership<-Water_URL%>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)")%>%
  html_text()
ownership

max_day_use_monthly<-Water_URL%>%
  html_nodes("th~ td+ td")%>%
  html_text()
max_day_use_monthly
max_day_use_monthly<-as.numeric(max_day_use_monthly)
class(max_day_use_monthly)
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
#making a df
water_df<-data.frame(water_system_name,
            PWSID,
            "Ownership"=ownership,
            "Max_day_use_monthly"=max_day_use_monthly)

#adding month & date columns
month<- c(1,5,9,2,6,10,3,7,11,4,8,12)
year<-c(2022)

water_df['Month']=month
water_df

water_df['Date']= my(paste(month, year))

water_df

#reordering for chronological order
water_chronological_df<- water_df[order(water_df$Month),]
water_chronological_df

#5 
ggplot(aes(x=Date, y= Max_day_use_monthly), data=water_chronological_df)+geom_point()+geom_line()+
  labs(title="Max Day Use for Durham in 2022", y="Max Day Use (MGD)")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 

scrape.it<-function(the_year, the_facility){
  base_URL<-'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
  the_scrape_url <- paste0(base_URL,'pwsid=',the_facility,'&year=', the_year)
  print(the_scrape_url)
  the_facility<-'03-32-010'
  the_website <- read_html(the_scrape_url)
  
  the_month<- c(1,5,9,2,6,10,3,7,11,4,8,12)
  
  water_system_name_anyyear<-the_website%>%
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)")%>%
    html_text()
  water_system_name
  
  PWSID_anyyear<-the_website%>%
    html_nodes("td tr:nth-child(1) td:nth-child(5)")%>%
    html_text()
  PWSID
  
  ownership_anyyear<-the_website%>%
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)")%>%
    html_text()
  ownership
  
  max_day_use_monthly_anyyear<-the_website%>%
    html_nodes("th~ td+ td")%>%
    html_text()
  
  df_water<- data.frame("Month" = the_month,
                               "Year" = the_year,
                               "Max_Day_Use" = as.numeric(max_day_use_monthly_anyyear)) %>% 
    mutate(Water_System_Name = !!water_system_name_anyyear,
           PWSID = !!PWSID_anyyear,
           Ownership = !!ownership_anyyear,
           Date = my(paste(Month,"-",Year)))
  return(df_water)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
max_daily_withdrawals_Durham_2015_df <- scrape.it(2015,'03-32-010')
max_daily_withdrawals_Durham_2015_df

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
max_daily_withdrawals_Asheville_2015_df <- scrape.it(2015,'01-11-010')
max_daily_withdrawals_Asheville_2015_df

max_daily_withdrawals_Asheville_Durham_2015_df<-
  bind_rows(max_daily_withdrawals_Asheville_2015_df,
            max_daily_withdrawals_Durham_2015_df)

max_daily_withdrawals_Asheville_Durham_2015_df

ggplot(max_daily_withdrawals_Asheville_Durham_2015_df, aes(x=Date, 
                                                           y=Max_Day_Use, color=Water_System_Name))+
  geom_point()+geom_line()+
  labs(title="Max Day Use for Asheville vs. Durham", y="Max Day Use (MGD)")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
#Mapping out the data from 2010-2021 for Asheville (Quicker way)
water_df_2010_to_2021_v2<-map2(seq(2010,2021), '01-11-010', scrape.it)
water_df_2010_to_2021_v2<-bind_rows(water_df_2010_to_2021_v2)

ggplot(water_df_2010_to_2021_v2,aes(y = Max_Day_Use, x=Date)) +
  geom_line()+geom_smooth(method="loess")+
  labs(title="Max Day Use for Asheville from 2010 to 2021", y="Max Day Use (MGD)")

#Manual way of mapping the data (feel free to ignore)
Asheville_2010 <- scrape.it(2010,'01-11-010')
Asheville_2011 <- scrape.it(2011,'01-11-010')
Asheville_2012 <- scrape.it(2012,'01-11-010')
Asheville_2013 <- scrape.it(2013,'01-11-010')
Asheville_2014 <- scrape.it(2014,'01-11-010')
Asheville_2015 <- scrape.it(2015,'01-11-010')
Asheville_2016 <- scrape.it(2016,'01-11-010')
Asheville_2017 <- scrape.it(2017,'01-11-010')
Asheville_2018 <- scrape.it(2018,'01-11-010')
Asheville_2019 <- scrape.it(2019,'01-11-010')
Asheville_2020 <- scrape.it(2020,'01-11-010')
Asheville_2021 <- scrape.it(2021,'01-11-010')

water_df_2010_to_2021_manual<-bind_rows(Asheville_2010, Asheville_2011,Asheville_2012, Asheville_2013, 
                                        Asheville_2014, Asheville_2015, Asheville_2016, Asheville_2017, 
                                        Asheville_2018, 
                                        Asheville_2019, Asheville_2020, Asheville_2021)

#Plot
ggplot(water_df_2010_to_2021_manual,
       aes(y = Max_Day_Use, x=Date)) +
  geom_line()+geom_smooth(method="loess")+
  labs(title="Max Day Use for Asheville from 2010 to 2021", y="Max Day Use (MGD)")

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
> Looking at the plot, the water usage in Asheville had a slight decline from 2010 to 2015 then increased from 2015 to 2021. 
