---
title: "Assignment 3: Data Exploration"
author: "Wynona Curaming"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packages
library(tidyverse, lubridate)

# 3. Import datasets
Neonics <- read.csv("~/ENV 872/EDA_Spring2024/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",
                    stringsAsFactors=TRUE)
Litter <- read.csv("~/ENV 872/EDA_Spring2024/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",
                   stringsAsFactors = TRUE)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids may affect the health of key insect pollinators like bees. Decline in bee populations will in turn have significant negative consequences on crop yields, which would then impact food security. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris play a role in carbon budgets and nutrient cycling (especially nitrogen, which is accumulated as they decompose). They also contribute to forest biodiversity by creating habitat for some animals like litter beetles.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Sampling is executed at terrestrial NEON sites that contain woody vegetation >2m tall. 
 2. Trap placement within plots may be either targeted or randomized, depending on the vegetation
 3. Over time some sampling plots may become impossible to sample, due to disturbance or other local changes

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
colnames(Neonics)
str(Neonics)
dim(Neonics)
```
> Answer: This dataset has 4623 observations and 30 variables. 

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer:The 5 most common effects studied in descending order are population, mortality, behavior, feeding behavior and reproduction. I think these would be of interest because we'd want to know how the chemical in this pesticide affects insect populations, such as whether there is an increase in mortality, and a difference to behavior (eg. feeding, reproduction). 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
sort(summary(Neonics$Species.Common.Name))
```

> Answer: The six most commonly studiesd species are the honey bee, parasitic wasp, buff tailed bumblebee, carniolan honey bee, bumble bee and italian honeybee. Bees are of interest over other insects because they are important pollinators, which are crucial for fruit formation in plants. On the other hand, the parasitic wasp is an issue to farmers because they lay their eggs in other insects. Once the larvae hatches, it kills the host insect. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
summary(Neonics$Conc.1..Author.)
```

> Answer: It is a factor. It is not numeric because some of the values have characters like "/" and "NR". We may need to do some data cleaning first, consulting domain-specific guidance. The NEON_Litterfall_UserGuide.pdf document doesn't seem to have information on this.   

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
studies_by_pub_year<-ggplot(Neonics) + 
  geom_freqpoly(aes(x=Publication.Year), binwidth=1) + 
  labs(title = "Number of Studies by Publication Year",
       x = "Publication Year",
       y = "Number of Studies")
studies_by_pub_year
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x=Publication.Year, color=Test.Location), binwidth=1)+
  labs(title = "Number of Studies by Publication Year",
       x = "Publication Year",
       y = "Number of Studies")
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:Overall, most common test location is on the lab, followed by the field (natural).From the mid-1990s to 2000 there were more studies based on the field (natural). From 2009 onwards (except for 2017), there were more lab studies than field studies. This probably shows a shift in disciplinary norms about what kind of test location is preferable.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics)+geom_bar(aes(x=Endpoint)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer: The two most common end points are NOEL and LOEL. NOEL refers to "No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls". LOEL refers to "Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls". 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
summary(Litter$collectDate)
dates_converted<-as.Date(Litter$collectDate)
class((dates_converted))
unique(dates_converted)
```
> Answer: The litter was sampled on August 2, 2018 and August 30, 2018. 

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
summary(Litter$namedLocation)
unique(Litter$namedLocation)
```

> Answer: Summary shows the number of observations for each plot Unique doesn't show the number of observations for each plot but only lists out the types of plots and number of types of plots.  

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter)+geom_bar(aes(x=functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter)+geom_boxplot(aes(x=functionalGroup, y=dryMass))
ggplot(Litter)+geom_violin(aes(x=functionalGroup, y=dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: There is not much variability in distribution within each functional group of litter. Hence, we can see only straight vertical or horizontal lines in the violin plot. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles tend to have the highest biomass at these sites. 
