---
title: "Assignment 2: Coding Basics"
author: "Wynona Curaming"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
sequencefrom1to30<-seq(1:30)
sequencefrom1to30
# I have assigned the sequence to an object. 

#2. 
meanseq <-mean(sequencefrom1to30)
medianseq <- median(sequencefrom1to30)
meanseq
medianseq

#3. 
mean(sequencefrom1to30)>median(sequencefrom1to30)

# I am calculating the mean of a sequence from 1 to 30 by creating a function
# Then I'm comparing it with the median of the same sequence by creating a function. 
# When running the code, we find that the mean is not larger than the median. 
# In this case, the mean and median are equal. 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

student_names <- c("Ariane","Nikhil","Anais", "Wynona") #character vector
test_score <- c(90, 86, 99, 49) #numeric vector
pass_status <- test_score>=50 #logical vector

# alternative method: class_performance<- cbind(student_names, test_score, pass_status)
class_performance <- data.frame (
  "Name" = student_names,
  "Score" = test_score,
  "Passed" = pass_status
)
class_performance
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A matrix would have the same data type across each column and no column names. Data frame could have columns with different data types and columns would have names.  

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.
```{r}
student_names <- c("Ariane","Nikhil","Anais", "Wynona") #character vector
test_score <- c(90, 86, 99, 49) #numeric vector
passing_status_2 <- ifelse(test_score >= 50, TRUE, FALSE)
  
class_performance3 <- data.frame (
  "Name" = student_names,
  "Score" = test_score,
  "Passed" = passing_status_2
)
return(class_performance3)
  
```


12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' worked because it generates a vector that can be combined with other vectors to create a data frame. 'if' and 'else' generate only functions based on conditional arguments. 
